---
title: "Review + recap <br> `r emo::ji('billed_cap')`"
author: ""
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    nature:
      highlightLines: true
      highlightStyle: solarized-dark
      countIncrementalSlides: false
      mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(dsbox)
```

## Announcements

- MT 01 due Tuesday


---

class: center, middle

# group_by

---

## What does group_by() do?

`group_by()` takes an existing `tbl` and converts it into a grouped `tbl` where operations are performed "by group":

.pull-left[
```{r}
ucbadmit
```
]
.pull-right[
```{r}
ucbadmit %>% 
  group_by(gender)
```
]

---

## What does group_by() not do?

`group_by()` does not sort the data, `arrange()` does:

.pull-left[
```{r}
ucbadmit %>% 
  group_by(gender)
```
]
.pull-right[
```{r}
ucbadmit %>% 
  arrange(gender)
```
]

---

## What does group_by() not do?

`group_by()` does not create frequency tables, `count()` does:

.pull-left[
```{r}
ucbadmit %>% 
  group_by(gender)
```
]
.pull-right[
```{r}
ucbadmit %>% 
  count(gender)
```
]

---

## Undo grouping with ungroup()

.pull-left[
```{r}
ucbadmit %>%
  count(gender, admit) %>%
  group_by(gender) %>%
  mutate(prop_admit = n / sum(n)) %>%
  select(gender, prop_admit)
```
]
.pull-right[
```{r}
ucbadmit %>%
  count(gender, admit) %>%
  group_by(gender) %>%
  mutate(prop_admit = n / sum(n)) %>%
  select(gender, prop_admit) %>%
  ungroup()
```
]

---

class: center, middle

# count

---

## count() is a short-hand

`count()` is a short-hand for `group_by()` and then `summarise()` to count the number of observations in each group:

.pull-left[
```{r}
ucbadmit %>%
  group_by(gender) %>%
  summarise(n = n()) 
```
]
.pull-right[
```{r}
ucbadmit %>%
  count(gender)
```
]

---

## count can take multiple arguments

.pull-left[
```{r}
ucbadmit %>%
  group_by(gender, admit) %>%
  summarise(n = n()) 
```
]
.pull-right[
```{r}
ucbadmit %>%
  count(gender, admit)
```
]

--

.question[
What is the difference between the two outputs?
]

--

- `count()` ungroups after itself
- `summarise()` peels off one layer of grouping
- The question mark just means that the number of groups is unkown right now, it will only be computed when/if the next line is executed

---

## tally() is also a short-hand

`tally()` is a short-hand for `summarise()`

.pull-left[
```{r}
ucbadmit %>%
  tally()
```
]
.pull-right[
```{r}
ucbadmit %>%
  summarise(n  = n())
```
]

--

<br>

.question[
What is the relationship between `count()` and `tally()`?
]

---

## Relationship between count() and tally()

`count()` is also a short-hand for `group_by()` and then `tally()`:

.pull-left[
```{r}
ucbadmit %>%
  group_by(admit) %>%
  tally()
```
]
.pull-right[
```{r}
ucbadmit %>%
  count(admit)
```
]

