---
title: "Implementing principles for effective	visualizations in R <br> `r emo::ji('woman_artist')`"
author: "Dr. Çetinkaya-Rundel"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    nature:
      highlightStyle: solarized-dark
      highlightLines: true
      countIncrementalSlides: false
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```


## Which is the best?

<iframe width="840" height="473" src="https://www.youtube.com/embed/AuJFuEq-qD8" frameborder="0" allow="autoplay; encrypted-media" allowfullscreen></iframe>

---

class: center, middle

# Principles	for	effective	visualizations

---

## Principles	for	effective	visualizations

- Order matters
- Put long categories on the y-axis
- Keep	scales consistent
- Select meaningful colors
- Use meaningful and nonredundant labels

---

## Data

A SurveyUSA poll<sup>✦</sup> asked 722 Los Angeles residents who described themselves as 
paying attention to the news about Brett Kavanaugh's confirmation to the Supreme 
Court of the United States the following question:

>Which best describes you:  
>I strongly support Kavanaugh's confirmation.  
>I support Kavanaugh's confirmation.  
>I oppose Kavanaugh's confirmation.  
>I strongly oppose Kavanaugh's confirmation.  

```{r echo=FALSE}
bk <- tibble(
  opinion = c(
    rep("Strongly support", 126), rep("Support", 147), rep("Oppose", 131), 
    rep("Strongly oppose", 215), rep("Not sure", 104)
    ),
  race = c(
    rep("White", 86), rep("Black", 3), rep("Hispanic", 26), rep("Asian", 11),
    rep("White", 93), rep("Black", 3), rep("Hispanic", 33), rep("Asian", 18),
    rep("White", 74), rep("Black", 7), rep("Hispanic", 28), rep("Asian", 22),
    rep("White", 129), rep("Black", 21), rep("Hispanic", 45), rep("Asian", 20),
    # with modification to make a point about consistent scales
    rep("White", 61), rep("Black", 0), rep("Hispanic", 21), rep("Asian", 22)
    )
)
```


.footnote[ 
<sup>✦</sup> Results of [SurveyUSA News Poll #24330](http://www.surveyusa.com/client/PollReport.aspx?g=b2c0e27f-02cc-4fde-8e6f-a026fe2f055f), retrieved Sep 18, 2018.
]

---

class: center, middle

# Order matters

---

## Alphabetical order rarely ideal

```{r}
ggplot(data = bk, aes(x = opinion)) +
  geom_bar()
```

---

## Use inherent level order

```{r}
bk <- bk %>%
  mutate(
    opinion = fct_relevel(opinion, "Strongly support", "Support", "Oppose", 
                                   "Strongly oppose", "Not sure")
  )

ggplot(data = bk, aes(x = opinion)) +
  geom_bar()
```

`fct_relevel`: Reorder factor levels using a custom order

---

## Alphabetical order rarely ideal

```{r}
ggplot(data = bk, aes(x = race)) +
  geom_bar()
```

---

## Order by frequency

```{r}
ggplot(data = bk, aes(x = fct_infreq(race))) +
  geom_bar()
```

`fct_infreq`: Reorder factors levels by frequency

---

## Clean up labels

```{r}
ggplot(data = bk, aes(x = fct_infreq(race))) +
  geom_bar() +
  labs(x = "Race", y = "Count")
```

---

class: center, middle

# Put long categories on 
# the y-axis

---

## Long categories can be hard to read

```{r}
ggplot(data = bk, aes(x = opinion)) +
  geom_bar()
```

---

## Move them to the y-axis

```{r}
ggplot(data = bk, aes(x = opinion)) +
  geom_bar() +
  coord_flip()
```

---

class: center, middle

# Pick a purpose

---

## Segmented bar plots can be hard to read

```{r}
ggplot(data = bk, aes(x = opinion, fill = race)) +
  geom_bar() +
  coord_flip()
```

---

## Use facets

```{r}
ggplot(data = bk, aes(x = opinion, fill = race)) +
  geom_bar() +
  coord_flip() +
  facet_grid(. ~ race)
```

---

## Avoid redundancy

```{r}
ggplot(data = bk, aes(x = opinion)) +
  geom_bar() +
  coord_flip() +
  facet_grid(. ~ race)
```

---

## Clean up labels

```{r}
ggplot(data = bk, aes(x = opinion)) +
  geom_bar() +
  coord_flip() +
  facet_grid(. ~ race) +
  labs(title = "Opinion on Brett Kavanaugh's confirmation", x = "", y = "")
```

---

class: center, middle

# Select meaningful colors

---

## Ordinal data aren't well represented by rainbow colors

```{r}
ggplot(data = bk, aes(x = race, fill = opinion)) +
  geom_bar(position = "fill") +
  coord_flip()
```

---

## Viridis scale works well with ordinal data 

```{r}
ggplot(data = bk, aes(x = race, fill = opinion)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_fill_viridis_d()
```

---

## Clean up labels

```{r}
ggplot(data = bk, aes(x = race, fill = opinion)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_fill_viridis_d() +
  labs(title = "Opinion on Brett Kavanaugh's confirmation", 
       subtitle = "by race", x = "", y = "", fill = "")
```

---

## Acknowledgements

These slides are based on earlier work by [Angela Zoss, Ph.D.](https://library.duke.edu/about/directory/staff/6881), Assessment & Data Visualization Analyst at Duke University Libraries.
