---
title: "Merge conflicts <br> `r emo::ji('fist_right')` `r emo::ji('fist_left')`"
author: "Dr. Çetinkaya-Rundel"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    nature:
      highlightStyle: solarized-dark
      highlightLines: true
      countIncrementalSlides: false
---

layout: true
  
<div class="my-footer">
<span>
Dr. Mine Çetinkaya-Rundel -
<a href="http://www2.stat.duke.edu/courses/Fall18/sta112.01/schedule" target="_blank">stat.duke.edu/courses/Fall18/sta112.01
</a>
</span>
</div> 

---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(emo)
library(dsbox)
library(datasauRus)
library(DT)
library(Tmisc)
```

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warn = 1
  )
# figure height, width, dpi
knitr::opts_chunk$set(fig.height = 2.5, fig.width = 5, dpi = 300) 
# fontawesome
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}
# code highlighting
hook_source <- knitr::knit_hooks$get('source')
knitr::knit_hooks$set(source = function(x, options) {
  x <- stringr::str_replace(x, "^[[:blank:]]?([^*].+?)[[:blank:]]*#<<[[:blank:]]*$", "*\\1")
  hook_source(x, options)
})
```


## Merge conflicts

- When two collaborators make changes to a file and push the file to their repo, git merges these two files.
- If these two files have conflicting content on the same line, git will produce a **merge conflict**.

---

## Getting started

- Clone your assignment repo in RStudio Cloud (`ae-05-dataviz-TEAMNAME`), and open the .Rmd file.
- Assign the numbers 1, 2, 3, and 4 to each of the team members.

---

## Let's cause a merge conflict

Take turns in completing the exercise, only one member at a time. For now, only commit and push the Rmd file:

- Member 1 - Change the team name to your actual team name, knit, commit, push.
- Member 2 - Change the team name to some other word, knit, commit, push. You should get an error. Pull. Take a look at the document with the merge conflict. Clear the merge conflict by choosing the correct/preferred change. Commit, and push.
- Member 3 - Add a name to the first code chunk, knit, commit, push. You should get an error. Pull. No merge conflicts should occur. Now push.
- Member 4 - Add a different name to the first code chunk, knit, commit, push. You should get an error. Pull. Take a look at the document with the merge conflict. Clear the merge conflict by choosing the correct/preferred change. Commit, and push.
- All members - Pull, and observe the changes in your document.

---

## Tips for collaborating via GitHub

- Always pull first before you start working.
- Commit, and push, often to minimize merge conflicts and/or to make merge conflicts easier to resolve.
- If you find yourself in a situation that is difficult to resolve, ask questions asap, don't let it linger and get bigger.
