---
title: "Multiple linear regression <br> `r emo::ji('woman_juggling')`"
author: ""
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    nature:
      highlightLines: true
      highlightStyle: solarized-dark
      countIncrementalSlides: false
      mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
---

```{r setup, child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
conflict_prefer("filter", "dplyr")
library(knitr)
library(broom)
library(DT)
library(plotly)
library(widgetframe)
```

## Announcements

- 

---

class: center, middle

# The linear model with multiple predictors

---

## Getting started

**Data:** Paris Paintings

```{r load-pp, message=FALSE}
pp <- read_csv("data/paris_paintings.csv", 
               na = c("n/a", "", "NA"))
```

---

## Multiple predictors

- Response variable: log(price)
- Explanatory variables: Width and height

```{r model-price-width-height}
m_wi_hgt <- lm(log(price) ~ Width_in + Height_in, data = pp)
tidy(m_wi_hgt)
```

--
- Linear model:
$$\widehat{log(price)} = 4.77 + 0.0269 width - 0.0133 height$$

---

## Visualizing models with multiple predictors

```{r plotly, echo=FALSE, warning=FALSE}
p <- plot_ly(pp, x = ~Width_in, y = ~Height_in, z = ~log(price),
        marker = list(size = 3,
                       color = "lightgray",
                       alpha = 0.5,
                       line = list(color = "gray",
                                   width = 2))) %>%
  add_markers() %>%
  plotly::layout(scene = list(xaxis = list(title = "Width (in)"),
                      yaxis = list(title = "Height (in)"),
                      zaxis = list(title = "Log(price)"))) %>%
  config(displayModeBar = FALSE)

frameWidget(p, width = "100%", height = "90%")
```

---

## Exploration 1

### Price, surface area, and living artist

- Explore the relationship between price of paintings and surface area, conditioned 
on whether or not the artist is still living
- First visualize and explore, then model

---

## Typical surface area

.question[
What is the typical surface area for paintings?
]

```{r viz-surf-artistliving, echo=FALSE,warning=FALSE}
ggplot(data = pp, 
       mapping = aes(y = log(price), x = Surface, color = factor(artistliving))) +
  geom_point(alpha = 0.3) +
  labs(color = "Living artist")
```

--

Less than 1000 square inches (which is roughly a painting that is 31in x 31in). There are very few paintings that have surface area above 5000 square inches.

---

## Narrowing the scope

For simplicity let's focus on the paintings with `Surface < 5000`:

```{r surf-lt-5000}
pp_Surf_lt_5000 <- pp %>%
  filter(Surface < 5000)
```

```{r viz-surf-lt-5000-artistliving, echo=FALSE, warning=FALSE}
ggplot(data = pp_Surf_lt_5000, 
       mapping = aes(y = log(price), x = Surface, color = factor(artistliving))) +
  geom_point(alpha = 0.3) +
  labs(color = "Living artist")
```

---

## Two ways to model

- **Main effects:** Assuming relationship between surface and logged price 
**does not vary** by whether or not the artist is living.
- **Interaction effects:** Assuming relationship between surface and logged 
price **varies** by whether or not the artist is living.

.pull-left[
```{r viz-main-effects, fig.height=4, echo=FALSE}
m_main <- lm(log(price) ~ Surface + factor(artistliving), data = pp_Surf_lt_5000)
m_main_aug <- augment(m_main)

ggplot(data = m_main_aug, 
       mapping = aes(y = log.price., x = Surface, color = factor.artistliving.)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = .fitted)) +
  labs(x = "Surface", y = "Log(price)", color = "Living artist", 
       title = "Main effects")
```
]
.pull-right[
```{r viz-interaction-effects, fig.height=4, echo=FALSE}
ggplot(data = pp_Surf_lt_5000,
       mapping = aes(y = log(price), x = Surface, 
                     color = factor(artistliving))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(y = "Log(price)", color = "Living artist", title = "Interaction effects")
```
]

---

## Fit model with main effects

- Response variable: log(price)
- Explanatory variables: Surface area and artist living (0/1 variable)

.midi[
```{r model-main-effects}
m_main <- lm(log(price) ~ Surface + factor(artistliving), 
             data = pp_Surf_lt_5000)
tidy(m_main)
```
]


--
- Linear model:
$$ \widehat{log(price)} = 4.88 + 0.000265~surface + 0.137~artistliving $$

---

## Solving the model

```{r viz-main-effects-2, fig.height=2, echo = FALSE}
ggplot(data = m_main_aug, 
       mapping = aes(y = log.price., x = Surface, color = factor.artistliving.)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = .fitted)) +
  labs(x = "Surface", y = "Log(price)", color = "Living artist")
```

--
- Non-living artist: Plug in 0 for `artistliving`
$\widehat{log(price)} = 4.88 + 0.000265~surface + 0.137 \times 0$  
$= 4.88 + 0.000265~surface$

--
- Living artist: Plug in 1 for `artistliving`
$\widehat{log(price)} = 4.88 + 0.000265~surface + 0.137 \times 1$  
$= 5.017 + 0.000265~surface$

---

## Visualizing main effects

```{r fig.height=2, echo = FALSE}
ggplot(data = m_main_aug, 
       mapping = aes(y = log.price., x = Surface, color = factor.artistliving.)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = .fitted)) +
  labs(x = "Surface", y = "Log(price)", color = "Living artist")
```

- **Same slope:** Rate of change in price as the surface area increases does 
not vary between paintings by living and non-living artists.
- **Different intercept:** Paintings by living artists are consistently more 
expensive than paintings by non-living artists.

---

## Interpreting main effects

.midi[
```{r exp-coefs}
tidy(m_main) %>% 
  mutate(exp_estimate = exp(estimate)) %>%
  select(term, estimate, exp_estimate)
```
]

- All else held constant, for each additional square inch in painting's surface area, the price of the painting is predicted, on average, to be higher by a factor of 1.
- All else held constant, paintings by a living artist are predicted, on average, to be higher by a factor of 1.15 compared to paintings by an artist who is no longer alive.
- Paintings that are by an artist who is not alive and that have a surface area of 0 square inches are predicted, on average, to be 132 livres.

---

## What went wrong?

.question[
Why is our linear regression model different from what we got from `geom_smooth(method = "lm")`?
]

.pull-left[
```{r echo=FALSE, fig.height=4}
ggplot(pp_Surf_lt_5000, aes(x = Surface, y = log(price), color = factor(artistliving))) + 
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(x = "Surface", y = "Log(price)", color = "Living artist")
```
]
.pull-right[
```{r echo=FALSE, fig.height=4}
m_pr <- lm(log(price) ~ Surface + factor(artistliving), data = pp_Surf_lt_5000)
m_pr_aug <- augment(m_pr)

ggplot(data = m_pr_aug, mapping = aes(y = log.price., x = Surface, color = factor.artistliving.)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = .fitted)) +
  labs(x = "Surface", y = "Log(price)", color = "Living artist")
```
]

---

## What went wrong? (cont.)

- The way we specified our model only lets `artistliving` affect the intercept.
- Model implicitly assumes that paintings with living and deceased artists have the *same slope* and only allows for *different intercepts*.  
- What seems more appropriate in this case? 
    + Same slope and same intercept for both colors
    + Same slope and different intercept for both colors
    + Different slope and different intercept for both colors?

---

## Interacting explanatory variables

- Including an interaction effect in the model allows for different slopes, i.e. 
nonparallel lines.
- This implies that the regression coefficient for an explanatory variable would 
change as another explanatory variable changes.
- This can be accomplished by adding an interaction variable: the product of two 
explanatory variables.

---

## Interaction: surface * artist living

.small[
```{r fig.height=2.5}
ggplot(data = pp_Surf_lt_5000,
       mapping = aes(y = log(price), x = Surface, 
                     color = factor(artistliving))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(x = "Surface", y = "Log(price)", color = "Living artist")
```
]

---

## Fit model with interaction effects

- Response variable: log(price)
- Explanatory variables: Surface area, artist living (0/1 variable), and 
their interaction

.midi[
```{r model-interaction-effects}
m_int <- lm(log(price) ~ Surface + factor(artistliving) + 
              Surface * factor(artistliving), 
            data = pp_Surf_lt_5000)
tidy(m_int)
```
]

- Linear model:
$$ \widehat{log(price)} = 4.91 + 0.00021~surface - 0.126~artistliving $$
$$+ ~ 0.00048~surface \times artistliving $$

---

## Interpretation of interaction effects

- Rate of change in price as the surface area of the painting increases does 
vary between paintings by living and non-living artists (different slopes), 
- Some paintings by living artists are more expensive than paintings by
non-living artists, and some are not (different intercept).

.small[
.pull-left[
- Non-living artist: 
$\widehat{log(price)} = 4.91 + 0.00021~surface$
$- 0.126 \times 0 + 0.00048~surface \times 0$
$= 4.91 + 0.00021~surface$
- Living artist: 
$\widehat{log(price)} = 4.91 + 0.00021~surface$
$- 0.126 \times 1 + 0.00048~surface \times 1$
$= 4.91 + 0.00021~surface$
$- 0.126 + 0.00048~surface$
$= 4.784 + 0.00069~surface$
]
.pull-right[
```{r fig.height=4, echo = FALSE}
ggplot(data = pp_Surf_lt_5000,
       aes(y = log(price), x = Surface, color = factor(artistliving))) +
  geom_point(alpha = 0.3) +
  stat_smooth(method = "lm", fullrange = TRUE)
```
]
]

---

## Third order interactions

- Can you? Yes
- Should you? Probably not if you want to interpret these interactions in context
of the data.

---

class: center, middle

# Quality of fit in MLR

---

## R-squared

- $R^2$ is the percentage of variability in the response variable explained by the 
regression model.

```{r}
glance(m_main)$r.squared
glance(m_int)$r.squared
```

--
- Clearly the model with interactions has a higher $R^2$.

--
- However using $R^2$ for model selection in models with multiple explanatory 
variables is not a good idea as $R^2$ increases when **any** variable is added 
to the model.

---

## R-squared - first principles

$$ R^2 =  \frac{ SS\_{Reg} }{ SS\_{Total} } = 1 - \left( \frac{ SS\_{Error} }{ SS\_{Total} } \right) $$

.question[
Calculate $R^2$ based on the output below.
]

```{r digits = 3}
anova(m_main)
```

---

## Adjusted R-squared

$$ R^2\_{adj} = 1 - \left( \frac{ SS\_{Error} }{ SS\_{Total} } \times \frac{n - 1}{n - k - 1} \right), $$

where $n$ is the number of cases and $k$ is the number of predictors in the model

--
- Adjusted $R^2$ doesn't increase if the new variable does not provide any new 
informaton or is completely unrelated.

--
- This makes adjusted $R^2$ a preferable metric for model selection in multiple
regression models.

---

## In pursuit of Occam's Razor

- Occam's Razor states that among competing hypotheses that predict equally well, 
the one with the fewest assumptions should be selected.
- Model selection follows this principle.
- We only want to add another variable to the model if the addition of that
variable brings something valuable in terms of predictive power to the model.
- In other words, we prefer the simplest best model, i.e. **parsimonious** model.

---

## Comparing models

It appears that adding the interaction actually increased adjusted $R^2$, so we 
should indeed use the model with the interactions.

```{r}
glance(m_main)$adj.r.squared
glance(m_int)$adj.r.squared
```

---

class: center, middle

# Model selection

---

## Backwards elimination

- Start with **full** model (including all candidate explanatory variables and 
all candidate interactions)
- Remove one variable at a time, and select the model with the highest adjusted 
$R^2$
- Continue until adjusted $R^2$ does not increase

---

## Forward selection

- Start with **empty** model
- Add one variable (or interaction effect) at a time, and select the model 
with the highest adjusted $R^2$
- Continue until adjusted $R^2$ does not increase

---

## Model selection and interaction effects

Model selection for models including interaction effects must follow the 
following two principles:

- If an interaction is included in the model, the main effects of both of 
those variables must also be in the model.
- If a main effect is not in the model, then its interaction should not be 
in the model.

---

## Other model selection criteria

- Adjusted $R^2$ is one model selection criterion
- There are others out there (many many others!), we'll discuss some later in 
the course, and some you might see in other courses

---

## Exploration 2

### Price, surface area, material, shape

.small[
```{r}
m_sms <- lm(log(price) ~ Surface + mat + Shape, data = pp)
tidy(m_sms) %>%
  print(n = 25)
```
]

---

## Shape and material

Collapse levels of `Shape` and `mat`erial variables with `forcats::fct_collapse`:

.small[
```{r}
pp <- pp %>%
  mutate(
    Shape = fct_collapse(Shape, oval = c("oval", "ovale"),
                                round = c("round", "ronde"),
                                squ_rect = "squ_rect",
                                other = c("octogon", "octagon", "miniature")),
    mat = fct_collapse(mat, metal = c("a", "br", "c"),
                            canvas = c("co", "t", "ta"),
                            paper = c("p", "ca"),
                            wood = "b",
                            other = c("e", "g", "h", "mi", "o", "pa", "v", "al", "ar", "m"))
  )
```
]

---

## Review fixes

.pull-left[
```{r}
pp %>%
  count(Shape)
```
]

.pull-right[
```{r}
pp %>%
  count(mat)
```
]

---

### Price, surface area, material, shape

```{r}
m_sms <- lm(log(price) ~ Surface + mat + Shape, data = pp)
tidy(m_sms)
```

---

## Application exercise: 

### `ae-10-model-selection-pp`

- First, plan the full model
- Then, fit, evaluate, select



